---
title: "Diversity and the Fossil Record"
output:
  html_document: default
  pdf_document: default
---
Written by J. N. Keating, 05/07/24

#### Getting started

First, lets install the required packages. 
```{r, message=FALSE}
if (!require(remotes)){ 
  install.packages("remotes") #Dev toolkit for installing R packages from github
}

if (!require(paleobioDB)){
  install.packages("paleobioDB") #package for extracting data from the PBDB
}
#or [remotes::install_github("ropensci/paleobioDB")] 

if (!require(palaeoverse)){
  install.packages("palaeoverse") #package with lots of useful functions for dealing with fossil time-series data
}
#or [remotes::install_github("palaeoverse/palaeoverse")]

if (!require(fossilbrush)){
  install.packages("fossilbrush") #package for cleaning up datasets from fossil databases e.g. the PBDB
}
#or [remotes::install_github("cran/fossilbrush")]

if (!require(divDyn)){
  install.packages("divDyn") #package for plotting diversity through time curves
}
#or [remotes::install_github("divDyn/r-package")]
```

Then load the libraries
```{r, message=FALSE}
library("paleobioDB")
library("fossilbrush")
library("palaeoverse")
library("divDyn")
```

#### Extracting, manipulating and plotting data from the PBDB

We can extract a dataset from the PBDB using the 'paleobioDB' package
```{r, message=FALSE}
dataset <- pbdb_occurrences(
    base_name = "cephalopoda",
    interval = c("Jurassic", "Neogene"),
    show = c("coords", "classext"),
    vocab = "pbdb",
    limit = "all"
)
head(dataset)
```

We can use fossilbrush to clean our dataset and fix taxonomy errors
```{r, message=FALSE}
b_ranks <- c("phylum", "class", "order", "family", "genus")
tax_check <- check_taxonomy(dataset, ranks = b_ranks, clean_name = T, resolve_duplicates = T)
dataset <- tax_check$data # replace raw dataset with cleaned dataset
```

Numeric ages may have changed since we downloaded the data from the PBDB. We can use look_up() to get up-to-date numeric ages for the occurrences from the International Commission on Stratigraphy (ICS).
```{r, message=FALSE}
# Get new numeric ages for named intervals using the interval key that is supplied with Palaeoverse
dataset <- look_up(dataset, int_key = interval_key)

# Make sure that any values which could not be matched contain their original values
dataset$interval_max_ma <- ifelse(is.na(dataset$interval_max_ma),
                                    dataset$max_ma,
                                    dataset$interval_max_ma)
dataset$interval_min_ma <- ifelse(is.na(dataset$interval_min_ma),
                                    dataset$min_ma,
                                    dataset$interval_min_ma)
dataset$interval_mid_ma <- (dataset$min_ma + dataset$max_ma)/2

# Rename columns so that new intervals are used for binning
PBDB_cols <- which(colnames(dataset) == "max_ma" | colnames(dataset) == "min_ma")
colnames(dataset)[PBDB_cols] <- c("PBDB_max_ma", "PBDB_min_ma")
GTS2020_cols <- which(colnames(dataset) == "interval_max_ma" | colnames(dataset) == "interval_min_ma")
colnames(dataset)[GTS2020_cols] <- c("max_ma", "min_ma")

# Identify outliers caused by mismatches between numeric ages and stratigraphic ages. Replace these with original ages from the PBDB.
out_min <- boxplot.stats(dataset$min_ma)$out
out_max <- boxplot.stats(dataset$max_ma)$out
out_min_ind <- which(dataset$min_ma %in% out_min)
out_max_ind <- which(dataset$max_ma %in% out_max)
dataset[out_max_ind, GTS2020_cols[[1]]] <- dataset[out_max_ind, PBDB_cols[[1]]]
dataset[out_min_ind, GTS2020_cols[[2]]] <- dataset[out_min_ind, PBDB_cols[[2]]]
dataset$interval_mid_ma <- (dataset$min_ma + dataset$max_ma)/2

# Look at the updated dataset
head(dataset[,c(8, 33:35)])
```

Fossilbrush can also help us identify anomalous ranges and occurrences. This is achieved b y constructing density profiles for taxon occurences and flagging those profiles with suspiciously long tails. 
```{r, message=FALSE}
pacm<-pacmacro_ranges(dataset,tail.flag=c(0.3,0.35,0.4), rank="genus",srt="max_ma",end="min_ma")
pranges <- pacm$kdensity
head(pranges[which(rowSums(pranges[,5:7]) == 3),]) #taxa with ranges flagged by fossilbrush
```

We can also visualise these anomalous ranges of any taxon and plot an interval spanning 95% of the occurrence density. 
```{r, message=FALSE}
dens<-densify(dataset)
plot_dprofile(dens,names(which(rowSums(pranges[,5:7]) == 3))[[1]],exit= FALSE)
abline(v= pacm$kdensity[names(which(rowSums(pranges[,5:7]) == 3))[[1]],"FAD95"],col="blue")
abline(v= pacm$kdensity[names(which(rowSums(pranges[,5:7]) == 3))[[1]],"LAD95"],col="blue")
```

In a real analysis it would be advisable to check each of the flagged taxa against the PBDB entries and published literature. For now, lets update our dataset using the 95% min and max ages of the flagged taxa. 
```{r, message=FALSE}
trimmed_ages <- dataset[,GTS2020_cols]
trimmed_ages[which(rowSums(pranges[,5:7]) == 3),] <- pranges[which(rowSums(pranges[,5:7]) == 3),c(3,4)]
colnames(trimmed_ages) <- c("max_ma", "min_ma")
trimmed_ages <- cbind(trimmed_ages, interval_mid_ma = (trimmed_ages$min_ma + trimmed_ages$max_ma)/2)
colnames(dataset)[seq(GTS2020_cols[1], GTS2020_cols[2])] <- c("GTS2020_max_ma", "GTS2020_interval_mid_ma", "GTS2020__min_ma")
dataset <- cbind(dataset, trimmed_ages)
```

In order to plot our time series data, we need to construct some time bins. One way to do this is to use stratigraphic intervals such as stages. A problem with this approach is that stratigraphic intervals have variable durations. The time_bins() function allows the user to specify a desired bin size, which is then used to create approximately equal-length time bins by grouping smaller intervals together.
```{r, message=FALSE}
bins <- time_bins(interval = c("Jurassic", "Neogene"),
                  rank = "stage",
                  scale = "GTS2020",
                  size = 5,
                  plot = FALSE)

head(bins)
```

Alternatively, we can specify equal time bins that are not constrained by stratigraphy. The number and size of our time bins will depend on the resolution of our occurrence data.  
```{r, message=FALSE}
# a simple function to create equal time bins
get_equal_bins <- function(min_ma, max_ma, n){
	bin_l <- (max_ma - min_ma)/n
	df <- data.frame(interval_name = paste("bin_", 1:n, sep = ""), max_ma = seq(from = min_ma + bin_l, to = max_ma, by = bin_l), min_ma = seq(from = min_ma, to =(max_ma - bin_l), by = bin_l))
	return(df)
}
eq_time_scale <- get_equal_bins(min_ma = 2.58, max_ma = 200, n = 40)
eq_bins <- time_bins(interval = c("Jurassic", "Neogene"),
                  rank = "stage",
                  scale = eq_time_scale,
                  plot = FALSE)

head(eq_bins)
```

Now lets bin the occurrences into our chosen time bins.
```{r, message=FALSE}
# Remove occurrences that are younger or older than the time intervals we're interested in
dataset <- subset(dataset, min_ma > min(bins$min_ma))
dataset <- subset(dataset, max_ma < max(bins$max_ma))

dataset <- bin_time(occdf = dataset,
                          bins = bins,
                          method = 'mid')
table(dataset$n_bins) # the number of occurrences per bin
```

We can use divDyn() to calculate a bunch of useful time-series metrics for analysing our data.
```{r, message=FALSE}
dd <- divDyn(dataset, bin="bin_midpoint", tax="genus")
head(dd)
```

Lets plot the Sampled-In-Bin diversity (the number of genera sampled in the focal bin), Boundary-Crosser diversity (the number of taxa with ranges crossing the boundaries of the interval) and the Range-Through diversity (taxa are considered to be present in all bins between the first and last appearance).
```{r, message=FALSE}
plot(dd$bin_midpoint, dd$divRT, # add the points
     axes = FALSE,
     xlim = rev(range(as.numeric(dd$bin_midpoint), na.rm = TRUE)), # reverse the x-axis
     ylab = "Richness (Diversity)", xlab = "",
     type = "l", lwd=2, col = "#0e826f") 
box()
axis(2) # add a normal axis to the elft side
axis_geo(side = 1, intervals = list("epochs", "periods"))
title(xlab = "Time (Ma)", line = 4)
lines(dd$bin_midpoint, dd$divBC, col="#F7CB15", lwd=2)
lines(dd$bin_midpoint, dd$divSIB, col="#F55D3E", lwd=2)
legend("topleft", legend=c("RT", "BC", "SIB"), col=c("#0e826f", "#F7CB15", "#F55D3E"), lwd=c(2,2,2), bg="white")
```

We can also pull out individual groups and compare their diversity curves. This might give us a better idea of how diversity changes through time.
```{r, message=FALSE}
d_amm <- subset(dataset, order == "Ammonitida")
d_nat <- subset(dataset, order == "Nautilida")
dd_amm <- divDyn(d_amm, bin="bin_midpoint", tax="genus")
dd_nat <- divDyn(d_nat, bin="bin_midpoint", tax="genus")
plot(dd_amm$bin_midpoint, dd_amm$divSIB, # add the points
     axes = FALSE,
     xlim = rev(range(as.numeric(dd$bin_midpoint), na.rm = TRUE)), # reverse the x-axis
     ylab = "Richness (Diversity)", xlab = "",
     type = "l", lwd=2, col = "#0e826f") 
box()
axis(2) # add a normal axis to the elft side
axis_geo(side = 1, intervals = list("epochs", "periods"))
title(xlab = "Time (Ma)", line = 4)
lines(dd_nat$bin_midpoint, dd_nat$divSIB, col="#F55D3E", lwd=2)
legend("topleft", legend=c("Ammonitida", "Nautilida"), col=c("#0e826f", "#F55D3E"), lwd=c(2,2), bg="white")
```

## Exercises
* Choose a taxonomic group and appropriate time interval. Extract the data from the PBDB, clean it and plot the Sampled-In-Bin diversity using equal time bins of 5 million years.  
